Notes on conformal manifolds

### 2025-05-20 Draft_MLFlatDirections.tex
Corrected small typos, added some comments and a reference.

		\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least fex points in each direction)}

@@

		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbors and perform a PCA on that local neighborhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.

@@

		\begin{figure}
			\centering
			\includegraphics[scale=0.5]{Figures/Local_PCA.png}
			\caption{Results of the local PCA analysis. The $x$-axis shows the dimension inferred by the algorithm, and the $y$-axis indicates the proportion of points for which that dimension was found. Each curve corresponds to a different value of $k$ in the $k$-nearest neighbors.}
			\label{local_pca}
		\end{figure}

@@

		Additionally, for $k \geq 20$, we observe an increasing number of points being assigned dimensionalities of 4 or 5. This behavior can be explained by the loss of locality when the number of neighbours becomes too large: increasing $k$ results in a coarser approximation, and the algorithm may then incorporate points that are no longer truly local. This artificial enlargement of the neighborhood can cause the estimated local dimensionality to rise.

@@

		\paragraph{Clusteting: HDBScan (only parameter: minimal number of points in a cluster), and graphs of 3d slices of the 5d space}

@@

		We set the minimum cluster size to 10 points. Out of the 986,060 3-dimensional points, the algorithm identifies three clusters: one of size 10, one of size 91,747, and one of size 6,849.

		The smallest cluster (10 points) is likely an artifact of local fluctuations in the data density and is not interpreted as physically meaningful. The last group, comprising 6,849 points, consists of points that the algorithm could not assign to any cluster. We interpret these as outliers or edge points rather than representatives of a separate manifold.

@@

		Now we have sampled the manifold and exctrated some basic information about it (namely its dimension and the fact that it is made of one block), we would like to see if we can exctract some analytic formula to caracterise it. What we have at the moment, are points on a 3-dimensional manifold, which are embedded in a 5-dimensional space. Therefore we can conclude that, in order to caracterise the manifold, we need two constraints on the embedding coordinates $\vec{x} = (x_1,x_2,x_4,x_8,x_{10})$. If we have a look at the form of the potential \eqref{eq:scalarpotential124810}, we see that if we use $\tilde{x_{8}} = e^{x_8}$, that this potential is actually a polynomial of the embedding coordinates. Therefore, the components of $|\nabla V|$ are also polynomials in those variables. We conclude that the constraints on the embedding coordinates we are looking for are polynomial constraints of the form $p(\vec{x}) = 0$, and that there are two of those. Of course if one takes directly the gradient of \eqref{eq:scalarpotential124810}, one ends up with such conditions, but none are usable directly to solve for two of the variables in terms of the others. The problem we are facing here is therefore a problem of symbolic regression : we are looking for analytic expressions (polynomial) that vanish once evaluated on our data points. There are a number of way one can deal with it, using already existing methods such as AI Feynman methods, or ..., or by doing rough gradient descent on the most general polynomial with arbitrary coefficients and trying to minimize its value once evaluated on the data points. In the spirit of trying to build a generalisable method, we develop here our own method, which is based on symbolic regression using an Annealing Importance Sampling method.

@@

		In the realm of symbolic regression, the objective is to uncover interpretable mathematical expressions that best describe a given dataset. In the problem discussed in this paper, we are looking for polynomial expressions such that $p(data) = 0$. This task involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. Markov Chain Monte Carlo (MCMC) techniques, while widely used, can struggle with poor mixing and getting trapped in local optima, especially in high-dimensional or multimodal spaces.

@@

		Let us now explain in more details how does this procedure goes. The goal is to reconstruct some distribution function $p(z)$, where here $z$ in going to be some polynomial. We will try to reconstruct this density function by a serie of density function $\pi_n(z_n) = \gamma_n(z_n)/Z_n$ with $n = 1,\dots,N$ is going to be the number of annealing steps, $\pi_n$ is defined in terms of an unnormalized density $\gamma_n$ and we have the normalizing constant $Z_n = \int \gamma_n(z) \mathrm{dz}$. We also assume we have a sequence of inverse temperature constants $\beta_n$, where $0 = \beta_1 < \beta_2 <\dots< \beta_N = 1$. We then define the unnormalized density at level $n$ in terms of a prior distribution $p_0(z)$ over the hypothesis space and an Loss function $L(z)$
		\begin{equation}
			\gamma_n(z) := p_0(z) \: \exp \left( -\beta_n L(z) \right)
		\end{equation}

@@

		Let's us now focus on the forward ($q(z_n^k | z_{n-1}^k)$) and backward propagation kernel ($q(z_{n-1}^k | z_n^k)$). The forward propagation kernel defines how we generate the state at time $n$ given the state at time $n-1$. It gives the probability to transition from $z_{n-1}$ to $z_n$. The backward kernel: It represents a hypothetical probability of transitioning back from state $z_n$ to state $z_{n-1}$. The purpose of $q$ is to give a proposition for $z_n$ given $z_{n-1}$. For the implementation of the two, we decide to implement an AIS-style MCMC code : we make some move in the space of polynomials, and then accept or reject those new polynomials based on an acceptance rate. So we first need to choose what are the available moves in the space of polynomials. Here are the choices we have made for our symbolic regression task : 
		\begin{itemize}
			\item Coefficient pertubation. Given a polynomial $z_{n-1}$, we choose randomly one of his coefficient and modify it by a random noise from the guassian distribution $\mathcal{N}(0,\sigma^2)$. So for eg : $2 x + 3 y^2 \rightarrow  2.1 x + 3 y^2$.
			\item Variable multiplication. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and multiply it by one of the available variable. So for eg : $2 x + 3 y^2 \rightarrow  2 x y + 3 y^2$.
			\item Variable division. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and divide it by one of its variable. So for eg : $2 x + 3 y^2 \rightarrow  2 x y + 3 y^2$.
		\end{itemize}
		This implementation allows us to straightforwardly calculate the forward and backward propagation kernel.  For example, the case of the coefficient pertubation is symmetric. Therefore $q(z_n^k | z_{n-1}^k) = q(z_{n-1}^k | z_n^k)$, which allows to simplify the calculation of the incremental importance weight.

>>>>

		\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}

@@

		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.

@@

		\begin{figure}
			\centering
			\includegraphics[scale=0.5]{Figures/Local_PCA.png}
			\caption{Results of the local PCA analysis. The $x$-axis shows the dimension inferred by the algorithm, and the $y$-axis indicates the proportion of points for which that dimension was found. Each curve corresponds to a different value of $k$ in the $k$-nearest neighbours.}
			\label{local_pca}
		\end{figure}

@@

		Additionally, for $k \geq 20$, we observe an increasing number of points being assigned dimensionalities of 4 or 5. This behavior can be explained by the loss of locality when the number of neighbours becomes too large: increasing $k$ results in a coarser approximation, and the algorithm may then incorporate points that are no longer truly local. This artificial enlargement of the neighbourhood can cause the estimated local dimensionality to rise.

@@

		\paragraph{Clustering: HDBScan (only parameter: minimal number of points in a cluster), and graphs of 3d slices of the 5d space}

@@

		We set the minimum cluster size to 10 points. Out of the 986,060 3-dimensional points, the algorithm identifies three clusters: one of size 10, one of size 91,747, and one of size 6,849. \ce{Are those numbers right? We loose more than 800,000 points?}

		The smallest cluster (10 points) is likely an artefact of local fluctuations in the data density and is not interpreted as physically meaningful. The last group, comprising 6,849 points, consists of points that the algorithm could not assign to any cluster. We interpret these as outliers or edge points rather than representatives of a separate manifold.

@@

		Now we have sampled the manifold and extracted some basic information about it (namely its dimension and the fact that it is made of one block), we would like to see if we can extract some analytic formula to characterise it. What we have at the moment, are points on a 3-dimensional manifold, which are embedded in a 5-dimensional space. Therefore we can conclude that, in order to characterise the manifold, we need two constraints on the embedding coordinates $\vec{x} = (x_1,x_2,x_4,x_8,x_{10})$. If we have a look at the form of the potential \eqref{eq:scalarpotential124810}, we see that if we use $\tilde{x}_{8} = e^{x_8}$, that, up to a global $e^{-2x_8}$ factor, this potential is actually a polynomial of the embedding coordinates. Therefore, the components of $|\nabla V|$ are also polynomials in those variables. We conclude that the constraints on the embedding coordinates we are looking for are polynomial constraints of the form $p(\vec{x}) = 0$, and that there are two of those. Of course if one takes directly the gradient of \eqref{eq:scalarpotential124810}, one ends up with such conditions, but none are usable directly to solve for two of the variables in terms of the others. The problem we are facing here is therefore a problem of symbolic regression: we are looking for analytic expressions (polynomial) that vanish once evaluated on our data points. There are a number of way one can deal with it, using already existing methods such as AI Feynman methods~\cite{Udrescu:2019mnk}, or ..., or by doing rough gradient descent on the most general polynomial with arbitrary coefficients and trying to minimize its value once evaluated on the data points. In the spirit of trying to build a generalisable method, we develop here our own method, which is based on symbolic regression using an Annealing Importance Sampling method.

@@

		In the realm of symbolic regression, the aim is to uncover interpretable mathematical expressions that best describe a given dataset. In the problem discussed in this paper, we are looking for polynomial expressions such that $p(data) = 0$. This task involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. Markov Chain Monte Carlo (MCMC) techniques, while widely used, can struggle with poor mixing and getting trapped in local optima, especially in high-dimensional or multimodal spaces.

@@

		Let us now explain in more details how does this procedure goes. The goal is to reconstruct some distribution function $p(z)$, where here $z$ in going to be some polynomial. \ce{Notational conflict: $p$ for polynomials and $p$ the distribution.} We will try to reconstruct this density function by series of density function $\pi_n(z_n) = \gamma_n(z_n)/Z_n$ with $n = 1,\dots,N$ is going to be the number of annealing steps, $\pi_n$ is defined in terms of an unnormalized density $\gamma_n$ and we have the normalizing constant $Z_n = \int \gamma_n(z) \mathrm{d}z$. We also assume we have a sequence of inverse temperature constants $\beta_n$, where $0 = \beta_1 < \beta_2 <\dots< \beta_N = 1$. We then define the unnormalized density at level $n$ in terms of a prior distribution $p_0(z)$ over the hypothesis space and a loss function $L(z)$:
		\begin{equation}
			\gamma_n(z) := p_0(z) \: \exp \left( -\beta_n L(z) \right).
		\end{equation}

@@

		Let's us now focus on the forward ($q(z_n^k | z_{n-1}^k)$) and backward propagation kernel ($q(z_{n-1}^k | z_n^k)$). The forward propagation kernel defines how we generate the state at time $n$ given the state at time $n-1$. It gives the probability to transition from $z_{n-1}$ to $z_n$. The backward kernel: It represents a hypothetical probability of transitioning back from state $z_n$ to state $z_{n-1}$. The purpose of $q$ is to give a proposition for $z_n$ given $z_{n-1}$. For the implementation of the two, we decide to implement an AIS-style MCMC code : we make some move in the space of polynomials, and then accept or reject those new polynomials based on an acceptance rate. So we first need to choose what are the available moves in the space of polynomials. Here are the choices we have made for our symbolic regression task : 
		\begin{itemize}
			\item Coefficient pertubation. Given a polynomial $z_{n-1}$, we choose randomly one of his coefficient and modify it by a random noise from the guassian distribution $\mathcal{N}(0,\sigma^2)$. So for eg : $2 x + 3 y^2 \rightarrow  2.1 x + 3 y^2$.
			\item Variable multiplication. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and multiply it by one of the available variable. So for eg : $2 x + 3 y^2 \rightarrow  2 x y + 3 y^2$.
			\item Variable division. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and divide it by one of its variable. So for eg : $2 x + 3 y^2 \rightarrow  2 x y + 3 y^2$. \ce{Change the example.}
		\end{itemize}
		This implementation allows us to straightforwardly calculate the forward and backward propagation kernel.  For example, the case of the coefficient pertubation is symmetric. Therefore $q(z_n^k | z_{n-1}^k) = q(z_{n-1}^k | z_n^k)$, which allows to simplify the calculation of the incremental importance weight.
